Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States.

### 2017-02-06 Manuscript/Manuscript_using_actual_data.Rmd
updated interpretation with new results

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

@@

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "stabbrs")], by = c("state"))
```

@@

diff <- bw_dat.males %>% arrange(era1.diff)
```

What we see so far (based on these 22 states):

@@

- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` (all other causes), 
`r diff[dim(diff)[1]-1, ]$state` (largely due to injuries), `r diff[dim(diff)[1]-2, ]$state` (injuries), `r diff[dim(diff)[1]-3, ]$state` (injuries), and `r diff[dim(diff)[1]-4, ]$state` (injuries).

- The gap increases for some states. Most notably in DC (CVD and other causes), Wisconsin (CVD), California (CVD). Also, for these states injuries do not contribute as much to narrowing the gap, so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.

- Injuries across the board contribute to narrowing the gap over time.

- Communicable disease narrows (except arizona), but much smaller contribution than injuries.

@@

- Generally, CVD increase gap.

- Cancer increases the gap everywhere. Size of contribution to change is a little smaller than CVD.

- Generally, non-communicable increase the gap but the contribution is tiny. Notable diff is in NYC and MD it decrease gap.

@@

- Except for Arizona and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, and CT. Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA. Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.

- Similar to the firts time frame, CVD and cancer also contributed to widening the gap over time. Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, and FL.

- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period.

@@

- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state `.

-By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state `.

@@

- Most states saw decreasing gaps in first era. Generally, all of the causes, except for cancer, were associated with changes that relatively benefitted blacks, with the biggest contributor to the narrowing gap being CVD and all other causes. However, in Wisconsin and CA, CVD led to a widening gap.

@@

- CVD differential by state. Contributes to widening the gap in CA and DC, for example, and narrowing it in KS and Kentucky.

- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit.

@@

>>>>

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

@@

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
```

@@

diff <- bw_dat.males %>% arrange(era1.diff)

max1 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983"])
mc1 <- cod.df$COD[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max1]

max2 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983"])
mc2 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max2]

max3 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983"])
mc3 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max3]

max4 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983"])
mc4 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max4]

max5 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983"])
mc5 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max5]
```

What we see so far (based on these `r dim(sum1)[1]` states):

@@

- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`, 
`r diff[dim(diff)[1]-1, ]$state` (`r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (`r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (`r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (`r round(max5,1)` due to `r mc5`).

- The gap increases for some states. Most notably in DC (CVD and other causes), Minnesota (CVD), Wisconsin (CVD), California (CVD). Also, for these states injuries do not contribute as much to narrowing the gap (and in Minnesota, injuries widens the gap), so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.

- Injuries across the board contribute to narrowing the gap over time, except for Minnesota.

- Communicable disease narrows (except Arizona), but much smaller contribution than injuries.

@@

- Generally, CVD increases gap.

- Cancer increases the gap everywhere except NM. Size of contribution to change is a little smaller than CVD.

- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, and Maryland where it decreases gap.

@@

- Except for New Mexico, Arizona, and Kentucky, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT. Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA. Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.

- Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time. Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, CT, Mississippi, and FL.

- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period, an increase of X, Y of which can be attributed to injuries.

@@

- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).

- By 2013, it ranges from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).

@@

- Most states saw decreasing gaps in first era.

- However, increases were found in Arizona, Oregon, Minnesota, and Wisconsin, largely due to cardiovascular disease. Thus, the effect of CVD on the LE varied by state over this time period in women.

- Otherwise, most of the other causes contributed to the gap narrowing, although cancer was associated with narrowing in some states (MS, NM, Oregon, for example) and increases in others (OH, Missouri, Kentucky).

- The biggest contributor to the narrowing gap being CVD (in most states, although heterogeniety) and all other causes.

@@

- CVD differential by state. Contributes to widening the gap in CA, Mississippi, and DC, for example, and narrowing it in Kansas, Delaware, and Kentucky.

- All other causes narrows across states (except DC, Indiana). Cancer increases the gap by a bit (except Oregon and Minnesota).

@@

- In a few states, Cancer contributed to widening the gap (NM, Oregon, Arizona, DC).
